---
title: "Unbiased Metabolomic Screen to identify biomarkers of drug response"
author: "Sony Tuteja, PharmD, MS"
output:
  html_document:
    depth: 3
    highlight: tango
    theme: paper
    toc: no
  pdf_document:
    toc: no
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
Niacin (vitamine B3) in large doses reduces plasma total cholesterol, low density lipoprotein cholesterol (LDL-C) and Triglycerides (TG). It can also raise high density lipoprotein cholesterol (HDL-C). However the mechanism by which niacin modulates plasma lipids is unknown. The dataset to be analyzed are results from an unbiased metabolomic profiling on plasma from healthy subjects receiving one dose of niacin. Supervised and unsupervised methods learned in this class will be applied to identify metabolites associated with lipid response to niacin.
https://github.com/sonytuteja/BMIN503_Final_Project 

### Introduction 
Atherosclerosis is the leading cause of death worldwide. Plasma lipids such as low-density lipoprotein cholesterol (LDL-C) and triglycerides (TG) are associated with coronary heart disease (CHD). Statins are the cornerstone of lipid lowering therapy reducing LDL-C by up to 50%, but even after aggressive LDL-C lowering there remains a high level of residual risk in susceptible individuals. There is a need for new hypolipidemic agents to modualate this risk. Niacin has been used for over 50 years to reduce plasma cholesterol, LDL-C and TGs. Niacin also raises high-density lipoprotein cholesterol (HDL-C), "the good cholesterol." However the mechanism by which niacin modulates plasma lipids is still unknown. I employed metabolomic profiling to identify metabolites associated with lipid response to niacin to identify drug-induced pertubations in metabolic pathways that are relevant to niacin's action on plasma lipids. The metabolomics dataset is very dense with hundreds of metabolites measured at each timepoint. I applied linear regression modeling to identify metabolites associated with niacin's effect on plasma lipids. In future analyses I will apply supervised and unsupervised analysis tools to identify significant metabolites that point to the mechanism of niacin.

The application of metabolomics to the discovery of biomarkers of drug response requires multidisciplanary expertise. It requires the technical knowledge of mass spectrometry assay development and metabolite identification. This experience will be useful to identify artifacts that may arise from the assay or sample preparation that may spuriously associated with the phenotype of interest. Expertise in clinical trial design is vital for appropriately designing the study to minimize confounders such as time of day or timing of meals around sample collection. Knowledge of pharmacology and metabolism is critical for data interpretation and follow up studies to infer biological relationships, for example off-target drug effects that are not relevant to the phenotype of interest but may be influencing metabolic pathway that are detected by metabolomics analysis. Lastly, expertise in analysis of multi-dimension data and statisics is critical for performing the correct normalization steps and methods to identify signficant features in the dataset.


### Methods
Pre- and post- niacin plasma samples at time 0,2,5,8 hours were profiled in 70 healthy subjects receiving 1 gm of niacin under fasting conditions using an untargeted gas-chromatography- mass spectrometry (GCMS) based metabolomics (Fiehn Lab, UC Davis, CA). The GCMS platform measured a total of 395 metabolites (180 known and 215 unknown). Complete data were returned for 67 subjects. Data were normalized by median and autoscaled using R package MetaboanalystR (Xia Lab). PCA analysis was performed to indentify outliers in the metabolite dataset. One-way anova was performed to determine which metabolites were significantly changed from time 1 (pre-niacin), with false discovery rate of 5% applied to account for multiple testing.

Plasma triglycerides, HDL-C, and total cholesterol (TC) were measured at time 0 and 8 hours. LDL-C was determined by using the Friedwall equation. The delta lipid values were determined by subtracting lipid levels at 0 hour from Lipid levels at 8 hr.

Linear regression models were built to determine which metabolites were significant predictors of delta TG, HDL-C, TC and LDL-C, including age, sex, race and body mass index as covariates using stepwise regression with backward elimination.

```{r eval=FALSE}
#install necessary packages for metabolomics analysis MetaboanalystR
metanr_packages <- function(){
  
  metr_pkgs <- c("Rserve", "RColorBrewer", "xtable", "som", "ROCR", "RJSONIO", "gplots", "e1071", "caTools", "igraph", "randomForest", "Cairo", "pls", "pheatmap", "lattice", "rmarkdown", "knitr", "data.table", "pROC", "Rcpp", "caret", "ellipse",
                 "scatterplot3d", "impute", "pcaMethods", "siggenes", "globaltest", "GlobalAncova", "Rgraphviz", "KEGGgraph", "preprocessCore", "genefilter", "SSPA", "sva")
  
  list_installed <- installed.packages()
  
  new_pkgs <- subset(metr_pkgs, !(metr_pkgs %in% list_installed[, "Package"]))
  
  if(length(new_pkgs)!=0){
    
    source("https://bioconductor.org/biocLite.R")
    biocLite(new_pkgs, dependencies = TRUE, ask = FALSE)
    print(c(new_pkgs, " packages added..."))
  }
  
  if((length(new_pkgs)<1)){
    print("No new packages added...")
  }
}
metanr_packages()

# Step 1: Install devtools
install.packages("devtools")
library(devtools)

# Step 2: Install MetaboAnalystR without documentation
devtools::install_github("xia-lab/MetaboAnalystR")

library(MetaboAnalystR)

```


```{r}
### Results

# Create object mset for storing processed data, read in data
mSet<-InitDataObjects("conc", "ts", FALSE)
mSet<-SetDesignType(mSet, "time")
mSet<-Read.TextData(mSet,"GENE_time_sex2.csv", "rowts", "disc")
mSet<-SanityCheckData(mSet)
mSet<-ReplaceMin(mSet)

#Normalization procedures- normalization by median and data scaling- mean-centered and divided by the standard deviation of each variable
mSet<-Normalization (mSet, "MedianNorm", "NULL", "AutoNorm", "GENE1003_D08", ratio=FALSE, ratioNum=20)

mSet<-PlotNormSummary(mSet, "norm_0_", "png", 72, width=NA)
mSet<-PlotSampleNormSummary(mSet, "snorm_0_", "png", 72, width=NA)

mSet<-PlotNormSummary(mSet, "norm_1_", "png", 72, width=NA)
mSet<-PlotSampleNormSummary(mSet, "snorm_1_", "png", 72, width=NA)
mSet<-SaveTransformedData(mSet)
```
Results of normalization procedure. See github for figures.

Figure 1-2. Metabolite normalization "norm_0_dpi72.png" and "norm_1_dpi72.png"
Figure 3-4. Sample normalization "snorm_0_dpi72.png" and "snorm_1_dpi72.png"


```{r eval=TRUE}

# Interactive PCA Plot (from MetaboanalystR). Viewed json file with LD Viz and took screen shots of PCA plot; see figure PCA_sex.png and PCA_time.png
mSet<-iPCA.Anal(mSet, "ipca_3d_0_.json")

# Perform ANOVA2 Analysis
mSet.anova<-ANOVA2.Anal(mSet, thresh=0.05, p.cor="fdr", type="w")

#subset metabolites acheiving significance
sigdatnorm <- read.csv("sigdatnorm.csv", header=TRUE)
str(sigdatnorm)


```
PCA plots with time and sex as factors- see github
Figure 5. PCA_sex.png
Figure 6. PCA_time.png

Table 1. Results of time series anova ranked by descending p-value- see github
67 metabolites were different by time, 32 metabolites different by sex and 2 significant by interaction. Table name anova_between_sbj.csv
```{r eval=TRUE}
#import phenotypes 
pheno <- read.csv("GENE_phenotypes.csv", header=TRUE)
str(pheno)
summary (pheno)
#look at distribution of major phenotype variables, the delta lipid values are normally  distributed.
library(ggplot2)
ggplot(data=pheno, aes(dhdl)) + 
  geom_histogram ()
ggplot(data=pheno, aes(dChol)) + 
  geom_histogram ()
ggplot (data=pheno, aes(dLDL)) +
  geom_histogram()
ggplot (data = pheno, aes(dTG)) +
  geom_histogram()
```

Results of linear regression models to predict delta HDL, LDL, TC, and TG.
```{r}
#first examine which clinical covariates predict change in lipid values 
summary (dHDL.lm <- lm(dhdl~ Sex + Race + Age + BMI + hdl_d7, data=pheno))
#sex and baseline HDL significantly predict the change in HDL
summary (dLDL.lm <- lm(dLDL~ Sex + Race + Age + BMI + ldl_d7, data=pheno))
#baseline LDL significantly predict the change in LDL
summary (dChol.lm <- lm(dChol~ Sex + Race + Age + BMI + chol_d7, data=pheno))
#baseline total Cholesterol significantly predict the change in Total Cholesterol
summary (dTG.lm <- lm(dTG~ Sex + Race + Age + BMI + tg_d7, data=pheno))
#baseline TG significantly predict the change in TG
```
```{r}
#subset time 1 datapoint (aka baseline signmetabolite values), merge to phenotype file
sigdatnormtime1 <- read.csv("sigdatnorm_time1.csv", header=TRUE)
metpheno <- merge(pheno, sigdatnormtime1, by=c("Studyid"), all=TRUE)
```

Results of linear regression models to predict delta HDL, LDL, TC, and TG including metabolites.
```{r}
#Linear Models for delta HDL
library(dplyr)
#subset variables pertinent to HDL analysis
hdl.data <-select(metpheno, -(flux_ind04:dChol))
hdl.data <- select(hdl.data, -(tg_d7:Sample))

#run linear model on half of the metabolites values because inputting all into a stepwise regression with backward elimination will not work
hdl.lmsub1<- lm(dhdl ~ .-(hdl_d18), data=hdl.data[,2:34]) 
summary (hdl.lmsub1)     
step(hdl.lmsub1, direction="backward")
#subset and run linear model on the other half of metabolites
hdlsub2.data <- data.frame(hdl.data[, 6:8], hdl.data[,35:74])
hdl.lmsub2<- lm(dhdl ~ .-(hdl_d18), data=hdlsub2.data)
summary (hdl.lmsub2)     
step(hdl.lmsub2, direction="backward")
#combine the top results from the sub models into one final model
hdl.lmcombo <- lm(dhdl ~ hdl_d7 + Sex + BMI + mannose + lysine + linolenic.acid + linoleic.acid + lauric.acid + lactic.acid + isoheptadecanoic.acid.NIST + indole.3.lactate +indole.3.acetate + ibuprofen + hypoxanthine + hydroxyproline.dipeptide.NIST +hexuronic.acid + glyceric.acid + glutamic.acid + glucose + fumaric.acid +docosahexaenoic.acid +capric.acid + aspartic.acid + aminomalonate + adipic.acid+ xylitol + valine + tyrosine +tryptophan + trans.4.hydroxyproline + threonine + stearic.acid + serine + palmitoleic.acid + palmitic.acid + oxoproline + ornithine + nicotinamide, data=hdl.data)
summary (hdl.lmcombo)
step(hdl.lmcombo, direction="backward")
step(dHDL.lm)

#with the clinical covariate model, sex and baseline hdl were significant predictors of delta HDL following niacin administration AIC=199.4. Including baseline metabolites values improved the prediction of delta HDL AIC=147.6. Final model waslm(formula = dhdl ~ hdl_d7 + Sex + mannose + linolenic.acid + inoleic.acid + lauric.acid + indole.3.lactate + indole.3.acetate + hypoxanthine + hexuronic.acid + glyceric.acid + glucose + fumaric.acid + capric.acid + aspartic.acid + adipic.acid + xylitol + valine + tyrosine + tryptophan + trans.4.hydroxyproline + threonine + stearic.acid + serine + palmitoleic.acid +palmitic.acid + oxoproline + ornithine + nicotinamide, data = hdl.data)

#adjustedr2 with this model was 0.7356 compared with 0.27 with the clinical variables alone, so including metabolites greatly improved variability in the change in HDL in response to niacin.
dhdl.lmcombo2<- lm(dhdl ~ hdl_d7 + Sex + mannose + linolenic.acid + linoleic.acid + lauric.acid + indole.3.lactate + indole.3.acetate + hypoxanthine + hexuronic.acid + glyceric.acid + glucose + fumaric.acid + capric.acid + aspartic.acid + adipic.acid + xylitol + valine + tyrosine + tryptophan + trans.4.hydroxyproline + threonine + stearic.acid + serine + palmitoleic.acid +palmitic.acid + oxoproline + ornithine + nicotinamide, data = hdl.data)
summary(dhdl.lmcombo2)

````



```{r}
#subset variables pertinent to LDL analysis
ldl.data <-select(metpheno, -(flux_ind04:dTG))
ldl.data <- select(ldl.data, -(apoc3_d7:Sample)) 

#run linear model on half of the metabolites values because inputting all into a stepwise regression with backward elimination will not work
ldl.lmsub1<- lm(dLDL ~ .-(ldl_d18), data=ldl.data[,2:34]) 
summary (ldl.lmsub1)     
step(ldl.lmsub1, direction="backward")
#subset and run linear model on the other half of metabolites
ldlsub2.data <- data.frame(ldl.data[, 6:8], ldl.data[,35:74])
ldl.lmsub2<- lm(dLDL ~ .-(ldl_d18), data=ldlsub2.data)
summary (ldl.lmsub2)     
step(ldl.lmsub2, direction="backward")
#combine the top results from the sub models into one final model
ldl.lmcombo <- lm(dLDL ~ Age + BMI +ldl_d7 + mannose + lysine + linolenic.acid +linoleic.acid +lauric.acid + lactic.acid + isonicotinic.acid +indole.3.lactate + hypoxanthine +hydroxyproline.dipeptide.NIST +histidine + hexuronic.acid + glycine + glycerol +glutamic.acid +erythritol + docosahexaenoic.acid + citrulline + capric.acid +behenic.acid + aspartic.acid + arachidonic.acid + adipic.acid +X3.hydroxybutyric.acid +xylitol +valine + tyrosine + tryptophan +trans.4.hydroxyproline +threonine + threonic.acid + threitol + succinic.acid + stearic.acid +serine + pyrophosphate + proline + phytanic.acid + phenylalanine +palmitoleic.acid + palmitic.acid + oxoproline + ornithine +oleic.acid + nicotinamide + N.methylalanine + N.acetylglycine + myristic.acid + methionine.sulfoxide + methionine+ X2.ketoglucose.dimethylacetal.NIST, data = ldl.data)
summary (ldl.lmcombo)
step(ldl.lmcombo, direction="backward")
step(dLDL.lm)
#with the clinical covariate model, BMI and baseline hdl were significant predictors of delta HDL following niacin administration AIC=334.28. Including baseline metabolites values improved the prediction of delta LDL AIC=192.67. Final model was Age + BMI + ldl_d7 + lysine + linolenic.acid + linoleic.acid + lactic.acid + isonicotinic.acid + indole.3.lactate + hypoxanthine + hydroxyproline.dipeptide.NIST + hexuronic.acid + glycine + glycerol + glutamic.acid + docosahexaenoic.acid + citrulline + capric.acid + behenic.acid + aspartic.acid + arachidonic.acid + adipic.acid + X3.hydroxybutyric.acid + tryptophan + threonine + threonic.acid + serine + proline + phytanic.acid + phenylalanine +     palmitoleic.acid + palmitic.acid + ornithine + oleic.acid + N.acetylglycine + methionine.sulfoxide + methionine + X2.ketoglucose.dimethylacetal.NIST 

#adjusted r2 with this model was 0.9057 compared with 0.16 with the clinical variables alone, so including metabolites greatly improved variability in the change in LDL in response to niacin.
dldl.lmcombo2<- lm(dLDL ~ ldl_d7 + Age + BMI + lysine + linolenic.acid + linoleic.acid + lactic.acid + isonicotinic.acid + indole.3.lactate + hypoxanthine + hydroxyproline.dipeptide.NIST + hexuronic.acid + glycine + glycerol + glutamic.acid + docosahexaenoic.acid + citrulline + capric.acid + behenic.acid + aspartic.acid + arachidonic.acid + adipic.acid + X3.hydroxybutyric.acid + tryptophan + threonine + threonic.acid + serine + proline + phytanic.acid + phenylalanine +palmitoleic.acid +palmitic.acid + ornithine + oleic.acid + N.acetylglycine + methionine.sulfoxide +methionine + X2.ketoglucose.dimethylacetal.NIST, data=ldl.data) 
summary(dldl.lmcombo2)

#removing nonsignificant features reduced the adjusted r2 to 0.881
dldl.lmcombo3<- lm(dLDL ~ ldl_d7 + Age + BMI + lysine + linolenic.acid + linoleic.acid + lactic.acid + indole.3.lactate + hypoxanthine + hydroxyproline.dipeptide.NIST + hexuronic.acid + glycine + glycerol + docosahexaenoic.acid + citrulline + behenic.acid + aspartic.acid + arachidonic.acid + adipic.acid + X3.hydroxybutyric.acid + tryptophan + serine + proline + phytanic.acid + phenylalanine +oleic.acid + N.acetylglycine +X2.ketoglucose.dimethylacetal.NIST, data=ldl.data) 
summary(dldl.lmcombo3)
```
```


Results
Acute niacin administration significantly perturbed plasma levels of 67 metabolites as shown by Table.anova_between_sbj.csv and Figure aov_2_dpi72.pdf. Clinical variables such as sex, BMI, age and race explain only very little of the variability in delta HDL, LDL, total cholesterol and triglycerides as shown in linear models (dHDL.lm, dLDL.lm, dChol.lm, dTG.lm). Including baseline levels of metabolites significantly changed by niacin in the regression models improved the amount of variability explained in the delta HDl and LDL models. The adjusted r2 for dHDL increased from 0.27 in the clinical covariates only model to 0.73 in the clinical covariate plus metabolites model. Similarly the amount of variation explained in the delta LDL increased from 0.16 with clinical covariates only to 0.91 when included baseline metabolites values. 

Conclusions
Baseline levels of circulating plasma metabolites can aid in predicting response to the HDL and LDL response to niacin treatment. Future efforts will focus on building these models for total cholesterol and triglycerides. In addition, machine learning approaches will be utilized to improved predictability of lipid response to niacin.